A Plasma Torus Around a Young Low-Mass Star

### 2025-03-29 paper/ms.tex
update

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  However, the
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
coherent sinusoidal Balmer emission at twice to four times the
star's equatorial velocity, directly demonstrating for the first
time the existence of corotating circumstellar plasma around a CPV.
Given that long-lived condensations of cool ($10^4$ K) plasma can
persist in the hot ($10^6$ K) coronae of stars with a wide range of
masses \cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can sustain an optically thick component; the origin
of the optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

@@

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  While Figure~\ref{fig:lc} highlights the main
object of interest in this article, over one hundred analogous objects
have now been found by K2 and TESS
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are phenomenologically identified based on their
structured, periodic optical light curves; most are M dwarfs with
rotation periods shorter than two days.  Within current sensitivity
limits, none have primordial disks \cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show this
behavior, an  observed fraction which decreases to $\approx$0.3\% by
$\approx$150\,Myr \cite{Rebull2020}.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectra of stars
with a wide range of masses have shown periodic high-velocity
absorption and emission in Balmer lines such as H$\alpha$, interpreted
as long-lived, corotating clumps of cool plasma
\cite{CollierCameron1989,CollierCameron1992,Barnes2000,Donati2000,Dunstone2006,Skelly2008,Leitzinger2016,Cang2021}.
Such clumps are thought to be forced into corotation by the magnetic
field, and the exact geometry of where the plasma can accumulate is
dictated by the field's topology.  For instance, a tilted dipole field
tends to yield an accumulation surface of a warped torus
\cite{Townsend2005}, whereas in the limit of a single strong field
line, accumulation occurs at the line's apex, furthest from the star
\cite{Waugh2022}.  To date, none of these spectroscopic variables have
shown any photometric anomalies \cite{Bouma2024}, leaving open the
issue of whether they are related to CPVs.

@@

% ** keck/hires: emission beyond v/veq>1
The spectroscopy shows emission beyond the star's equatorial velocity
($v_{\rm eq}$=130\,\kms).  There are at least two distinct emission
components, separated by half a cycle in phase.  The first component
has clearer sinusoidal behaviour and is double-peaked, with peak
semi-amplitudes of $K_1$=2.1\,$v_{\rm eq}$ and 2.7\,$v_{\rm eq}$.
There is significant non-periodic variability in the emissivity of
this double-peaked component: the flux excess begins with an amplitude
of 70\% above the continuum early in the observation sequence, and
diminishes to 10\% by its end.  The component 180$^\circ$ opposite in
phase is detected only from $\phi$=0.2-1.0.  From $\phi$=0.2-0.5, this
latter component appears connected to the star in velocity space.
While its peak semi-amplitude of $K_1$=3.9\,$v_{\rm eq}$ is achieved
at both $\phi$=0.25 and 0.75, its amplitude similarly decreases from a
60\% excess over the continuum at the beginning of the observation
sequence to a 10\% excess by its end.  The apparent period for all
of these emission components is consistent with the photometric
\periodhr\ hour period.

These sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to corotate with the star.  Based on the
period, this material's motion is not Keplerian; it can only be
explained by plasma being dragged along with the rotating stellar
magnetic field.  The velocity semi-amplitude of the sinusoids gives
the distance of the clumps from the stellar surface:
2.1-2.7\,$R_\star$ for the closer clump, and 3.9\,$R_\star$ for the
other.   These clumps transit in front of the star when passing from
negative to positive velocity.  This implies that the spectroscopic
transit of the 2.1-2.7\,$R_\star$ clump occurs simultaneously with the
sharp photometric spike visible in the TESS data.

@@

Spectra of magnetically-active, rapidly rotating stars with a wide range
of masses have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} and
transient absorption features
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for
such spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar corona
that can last days to weeks \cite{VialEngvold2015}.  These
condensations fall back to the Sun's surface because gravity is
stronger than any magnetic tension or centrifugal force capable of
sustaining them.  However for stars with magnetospheric radii $R_{\rm
m}$ that exceed their corotation radii $R_{\rm c}$, the effective
potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to accumulate over much
longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit nor be opaque in
broadband optical light.
For instance, the H$\alpha$ emission clump at 3.9\,$R_\star$ shows no 
photometric signal in the February 2024 photometry.

@@

Regarding the upper limit, Extended Data Figure~\ref{fig:sed} shows the SED.  While
AllWISE \cite{Cutri2014} yielded a confident W3 detection
(9.8$\sigma$), the W4 extraction yielded only a marginal indication
(1.7$\sigma$) of detectable flux.  Similar to other CPVs
\cite{Stauffer2017,Bouma2024}, the photometric uncertainties from WISE
W1 and W2 allow at most a $\lesssim$2\% excess at 3-5\,$\mu$m relative
to the stellar photosphere, and a $\lesssim$5\% excess at 10\,$\mu$m
(W3).  To estimate the implied mass bound, we assume a dust
temperature $T_{\rm d}$=1500\,K, typical for dust near the star (see
\cite{Zhan2019} for discussion regarding dust sublimation).  We then
treat emission from the dust and star as Planck functions, and require
$L_{\rm d} < f L_\star$, where the factor $f$ is set by the
photometric precision of WISE and $L_{\rm d}$ is the bolometric dust
luminosity.  Given the reported uncertainties, we numerically find
$f<6\cdot$10$^{-3}$.  From the Stefan-Boltzmann law we can then write
$A_{\rm d} < f (T_\star / T_{\rm d})^4 Q_{\rm em}^{-1} (4\pi
R_\star)^2$, for $A_{\rm d}$ the total emitting surface area of the
dust, and $Q_{\rm em}$ an emission efficiency parameter.  Converting
this constraint to a dust mass requires an assumption regarding the
grain properties.  We assume a grain density $\rho_{\rm
d}$=3\,g\,cm$^{-3}$ typical for silicate grains, a fixed grain size
$a$=1\,$\mu$m, and no self-absorption.  This enables the assumption
that $A_{\rm d} = N \pi a^2$, for $N$ the total number of dust grains.
This in turn yields an upper limit on the dust mass of
\begin{equation}
  M_{\rm dust} \lesssim 4 \cdot 10^{17}\, {\rm g}\ 
  \left( \frac{f}{6\cdot10^{-3}} \right)
  \left( \frac{T_\star}{3000\,{\rm K}} \frac{1500\,{\rm K}}{T_{\rm d}} \right)^4
  \left( \frac{Q_{\rm em}}{1} \right)^{-1}
  \left( \frac{R_\star}{0.4\,R_\odot} \right)^2
  \left( \frac{a}{1\,\mathrm{\mu m}} \right)
  \left( \frac{\rho_{\rm d}}{3\,\mathrm{g\,cm^{-3}}} \right).
\end{equation}

@@

Three relevant objects for comparison to these limits are solar
prominences, planetesimals, and comets.  Prominences of the Sun have
gas masses of order $10^{14}$\,g-$10^{16}$\,g \cite{VialEngvold2015}.
A planetesimal of mass $\approx$10$^{15}$\,g with a bulk density of
1\,g\,cm$^{-3}$ would have a diameter of order one kilometer.
Halley's comet has a mass of order $10^{17}$\,g \cite{Rickman1989}, of
which $\sim$$10^{14}$\,g is shed per orbit, most of which inspirals
toward the Sun due to Poynting-Robertson drag.

These calculations suggest that if dust is responsible for the
broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.  Given
$M_{\rm gas}$$\approx$8$\times$10$^{16}$\,g from
Section~\ref{subsec:gas}, the allowed dust masses imply $M_{\rm
gas}/M_{\rm dust}$ ranges of 1-100.  More careful measurements of this
ratio---in particular enabled by inferring the dust mass through high
precision infrared spectrophotometry---could provide a path for
distinguishing the scenario of a trapped stellar outflow from an accumulation
of externally-sourced material.

>>>>

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  However, the
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
sinusoidal time-varying Balmer emission at twice to four times the
star's equatorial velocity, directly demonstrating for the first
time the existence of corotating circumstellar plasma around a CPV.
Given that long-lived condensations of cool ($10^4$ K) plasma can
persist in the hot ($10^6$ K) coronae of stars with a wide range of
masses \cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can sustain an optically thick component; the origin
of the optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

@@

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  While Figure~\ref{fig:lc} highlights the main
object of interest in this article, over one hundred analogous objects
have now been found by K2 and TESS
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are phenomenologically identified based on their
structured, periodic optical light curves; most are M dwarfs with
rotation periods shorter than two days.  Within current sensitivity
limits, none host disks \cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show complex
photometric behavior, an observed fraction which decreases to $\approx$0.3\% by
$\approx$150\,Myr \cite{Rebull2020}.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectra of stars
with a wide range of masses have shown periodic high-velocity
absorption and emission in Balmer lines such as H$\alpha$, interpreted
as long-lived, corotating clumps of cool plasma
\cite{CollierCameron1989,CollierCameron1992,Barnes2000,Donati2000,Dunstone2006,Skelly2008,Leitzinger2016,Cang2021}.
Such clumps are thought to be forced into corotation by the magnetic
field, and the exact geometry of where the plasma can accumulate is
dictated by the field's topology.  For instance, a magnetic dipole field
tilted with respect to the stellar spin axis
yields accumulations in a warped torus geometry
\cite{Townsend2005}, whereas in the limit of a single strong field
line, accumulation occurs at the line's apex, furthest from the star
\cite{Waugh2022}.  To date, none of these spectroscopic variables have
shown any photometric anomalies \cite{Bouma2024}, leaving open the
issue of whether they are related to CPVs.

@@

% ** keck/hires: emission beyond v/veq>1
The spectroscopy shows emission beyond the star's equatorial velocity
($v_{\rm eq}$=130\,\kms).  There are at least two distinct emission
components, separated by half a cycle in phase.  The first component
has clearer sinusoidal behaviour in time and is double-peaked, with peak
semi-amplitudes of $K_1$=2.1\,$v_{\rm eq}$ and 2.7\,$v_{\rm eq}$.
There is significant non-periodic variability in the emissivity of
this double-peaked component: the flux excess begins with an amplitude
of 70\% above the continuum early in the observation sequence, and
diminishes to 10\% by its end.  The component 180$^\circ$ opposite in
phase is detected only from $\phi$=0.2-1.0.  From $\phi$=0.2-0.5, this
latter component appears connected to the star in velocity space.
While its peak semi-amplitude of $K_1$=3.9\,$v_{\rm eq}$ is achieved
at both $\phi$=0.25 and 0.75, its amplitude similarly decreases from a
60\% excess over the continuum at the beginning of the observation
sequence to a 10\% excess by its end.  The apparent period for all
of these emission components is consistent with the photometric
\periodhr\ hour period.

These sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to corotate with the star.  Based on the
observed sinusoidal period and peak velocity, this material's motion
is not controlled by gravitational attraction to the star; it can only be
explained by plasma being dragged along with the rotating stellar
magnetic field.  The velocity semi-amplitude of the sinusoids gives
the distance of the clumps from the stellar surface:
2.1-2.7\,$R_\star$ for the closer clump, and 3.9\,$R_\star$ for the
other.   These clumps transit in front of the star when passing from
negative to positive velocity.  This implies that the spectroscopic
transit of the 2.1-2.7\,$R_\star$ clump occurs simultaneously with the
sharp photometric spike visible in the TESS data.

@@

Spectra of magnetically-active, rapidly rotating stars with a wide range
of masses have been known to exhibit both sinusoidal emission features
in time
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} and
transient absorption features
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for
such spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar corona
that can last days to weeks \cite{VialEngvold2015}.  These
condensations fall back to the Sun's surface because gravity is
stronger than any magnetic tension or centrifugal force capable of
sustaining them.  However for stars with magnetospheric radii $R_{\rm
m}$ that exceed their corotation radii $R_{\rm c}$, the effective
potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to accumulate over much
longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit nor be opaque in
broadband optical light.
For instance, the H$\alpha$ emission clump at 3.9\,$R_\star$ shows no 
photometric signal in the February 2024 photometry.

@@

Regarding the upper limit, Extended Data Figure~\ref{fig:sed} shows the SED.  While
AllWISE \cite{Cutri2014} yielded a confident W3 detection
(9.8$\sigma$) consistent with the photometric extrapolation from bluer
bandpasses, the W4 extraction yielded only a marginal indication
(1.7$\sigma$) of detectable flux.  Similar to other CPVs
\cite{Stauffer2017,Bouma2024}, the photometric uncertainties from WISE
W1 and W2 allow at most a $\lesssim$2\% excess at 3-5\,$\mu$m relative
to the stellar photosphere, and a $\lesssim$5\% excess at 10\,$\mu$m
(W3).  To estimate the implied mass bound, we assume a dust
temperature $T_{\rm d}$=1500\,K, typical for dust near the star (see
\cite{Zhan2019} for discussion regarding dust sublimation).  We then
treat emission from the dust and star as Planck functions, and require
$L_{\rm d} < f L_\star$, where the factor $f$ is set by the
photometric precision of WISE and $L_{\rm d}$ is the bolometric dust
luminosity.  Given the reported uncertainties, we numerically find
$f<6\cdot$10$^{-3}$.  From the Stefan-Boltzmann law we can then write
$A_{\rm d} < f (T_\star / T_{\rm d})^4 Q_{\rm em}^{-1} (4\pi
R_\star)^2$, for $A_{\rm d}$ the total emitting surface area of the
dust, and $Q_{\rm em}$ an emission efficiency parameter.  Converting
this constraint to a dust mass requires an assumption regarding the
grain properties.  We assume a grain density $\rho_{\rm
d}$=3\,g\,cm$^{-3}$ typical for silicate grains, a fixed grain size
$a$=1\,$\mu$m, and no self-absorption.  This enables the assumption
that $A_{\rm d} = N \pi a^2$, for $N$ the total number of dust grains.
This in turn yields an upper limit on the dust mass of
\begin{equation}
  M_{\rm dust} \lesssim 4 \cdot 10^{17}\, {\rm g}\ 
  \left( \frac{f}{6\cdot10^{-3}} \right)
  \left( \frac{T_\star}{3000\,{\rm K}} \frac{1500\,{\rm K}}{T_{\rm d}} \right)^4
  \left( \frac{Q_{\rm em}}{1} \right)^{-1}
  \left( \frac{R_\star}{0.4\,R_\odot} \right)^2
  \left( \frac{a}{1\,\mathrm{\mu m}} \right)
  \left( \frac{\rho_{\rm d}}{3\,\mathrm{g\,cm^{-3}}} \right).
\end{equation}

@@

Three relevant objects for comparison include solar
prominences, planetesimals, and comets.  Prominences of the Sun have
gas masses of $10^{14}$\,g-$10^{16}$\,g \cite{VialEngvold2015}.
A planetesimal of mass $\approx$10$^{15}$\,g with a bulk density of
1\,g\,cm$^{-3}$ would have a diameter of order one kilometer.
Halley's comet has a mass of order $10^{17}$\,g \cite{Rickman1989}, of
which $\sim$$10^{14}$\,g is shed per orbit, most of which inspirals
toward the Sun due to Poynting-Robertson drag.

Broadly, these considerations suggest that if dust is responsible for
the broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.  Given
$M_{\rm gas}$$\approx$8$\times$10$^{16}$\,g from
Section~\ref{subsec:gas}, the allowed dust masses imply $M_{\rm
gas}/M_{\rm dust}$ ranges of 1-100.  More careful measurements of this
ratio---in particular by inferring the dust mass through high
precision infrared spectrophotometry---could provide a path for
distinguishing the scenario of a trapped stellar outflow from an
accumulation of externally-sourced material.  While there are several
plausible external sources, feeding through a low-mass disk in
particular cannot be ruled out based on typical disk depletion times
\cite{Haisch2001}.  Observations of infrared excesses and accretion
signatures in low-mass stars tens of millions of years old suggest a
broad lifetime distribution for such disks
\cite{Silverberg2020,Lee2020,Gaidos2022,Pfalzner2024}.
